---
title: "Catch curve analysis"
author: "Matt Navarro"
date: "10/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

We are going to use a linear catch curve technique to determine whether a stock is being overfished. The data are simulated, but are based closely on the
WA Dhufish <http://www.fish.wa.gov.au/species/wa-dhufish/Pages/default.aspx>

This is a long lived endemic species in WA and is a key species in WA's recreaitonal and commercial fisheries. 

We are going to present analysis based on two data sets. They each contain age frequencies of 500 fish. 
The first is based on constant recruitment - this means that the same number of juvenile fish enter the fishery each year. The second is based on variable recruitment - his means that the number of juvenile fish entering the fishery varies by year

Variable recruitment is probably more realistic as recruitment for many fish species is driven by oceanographic features (e.g. currents). We will talk about this more in the third lecture. 

For each data set we want to know:

-Is the stock overfished? And,

-Is our analysis reliable? 

## Big thanks and a disclaimer

This script was prepared largely by Alex Hesp from WA Fisheries (Department of Primary Industries and Regional Development) just for you guys. Bit of a legend!

### Disclaimer:
The author does not warrent that the information provided is free from errors or omissions. 
The author do not accept any form of liability, for the information provided, or for any
consequences arising from its use or any reliance placed upon it. The information, opinions
and advice provided may not relate, or be relevant, to a reader's/user's particular circumstance. 
Readers must evaluate the relevance for their purposes of the provided information, and actively 
seek review from the Department prior to publishing or entering into any commitment based on the information.
Opinions expressed by the author are the individual opinions expressed by that person
and are not necessarily those of the Department.


## Main analysis
### Load the constant recruitment data and have a look
```{r}
data_const=read.csv("C:/GitHub/SCIE2204/Catch curve/AgeDatWithConstRec.csv", header=T, row.names = NULL)
plot(data_const$Age, data_const$Frequency, "o", main="Age sample data", cex.main=1.0, 
        xlab="Age class",ylab="Frequency", frame=F, ylim=c(0,80))
```

The sample consists of very few young fish. This probably means that it comes from a commercial or recreational fishery. 

Frequency peaks at 7 years at around 65 fish.it then trails off till aroud 20 and is relatively flat above age 20. 

### Setting up the data

For catch curve analysis we are only really interested in the ages that the fishery is impacting on. A common assumption is to use the peak age frequency + 1. looking at the plot we can see the peak frequency is 7 and so the cut off should be 8. 

```{r}
MinCutOff = data_const[which.max(data_const$Frequency),"Age"] + 1
MinCutOff

plot(data_const$Age, data_const$Frequency, "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F, ylim=c(0,80))
abline(v = MinCutOff, col="red", lty=2)
```
We are also not interested in really old fish that aren't necessarily kept in the fishery. We will, somewhat arbitrarily, say that we are interested in all ages up to the age where the frequency of fish is just 1. 

```{r}
MaxCutOff <- data_const[match( 1 , data_const$Frequency), "Age"]
plot(data_const$Age, data_const$Frequency, "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F, ylim=c(0,80))
abline(v = c(MinCutOff, MaxCutOff), col="red", lty=2)
```

So lets take the data sub-set and plot the data

```{r}
data_analysis<- data_const[ data_const$Age >= MinCutOff & data_const$Age <= MaxCutOff,  ]
plot(data_analysis$Age, data_analysis$Frequency, "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F)
```

### Catch curve analysis - estimating total mortality
Catch curve analysis involves fitting a straight line to the agre frequency data. Do you think that a straight line would fit this data well? I don't really think so - at age 20+ the data is more or less flat. 

One option to fix this is to take a log transformation of the Frequency. 

```{r}
plot(data_analysis$Age, log(data_analysis$Frequency), "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F)
```

That looks better!

Now lets fit a model. 

The code below fits a linear model to the log of frequency against Age for our data subset. Basically it fits a straight line through the above plot. 

```{r}
CatchCurveModel <- lm(log(Frequency) ~ Age, data = data_analysis)
```

We have fit a linear model to the log of frequency against Age for our data subset. Basically the model is fitting a straight line to the plot above. We can have a look at the line:

```{r}
EstlnFreq <- predict(lm(log(Frequency) ~ Age, data = data_analysis))
plot(data_analysis$Age, log(data_analysis$Frequency), "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F)
lines(data_analysis$Age,EstlnFreq,"l",col="blue")
```

Seems like a reasonable fit. 

Now, the whole reason we made this model was to estimate the total mortality (Z) in the population. We can do this by looking at how frequency drops off with age across our data subset - that is the slope of the blue line in our plot. 

We can have a look at the slope by looking at the model summary

```{r}
summary(CatchCurveModel)
Z <- CatchCurveModel$coefficients[[2]]*-1
Z
```
The slope, which is equal to the total mortality rate (Z), is next to Age. It is 0.23. 

### Is overfishing occuring?

You will recall from the lecture that a rule of thumb is that overfishing is not ocuring if the fishing mortality rate is less than the natural mortality rate. 
To start to determine if this is true we need to estimate the natural mortality rate. 

Natural mortality can be estimated using the maximum observed age for the fish species which is 40, and using the formula below based on the paper at: <https://paperpile.com/shared/GSKSGX>

```{r}
M <- exp(1.46 - 1.01*log(40))
M
```
So we now know that total mortality (Z) is 0.23 and natural mortality (M) is 0.10. 

All fish either die naturally or are caught, so we can estimate fishing mortality by F = Z - M

```{r}
F <- Z - M
F
```

Finally we can assess whether overfishing is occuring by examining whether F > M

```{r}
F > M
```
So F is 0.12 and M is 0.10, so it appears that overfishing is indeed occuring, or atleast accoding to this analysis. 

## variable recruitment data

Lets now look at the variable recrtuitment data. 
```{r}
data_var=read.csv("C:/GitHub/SCIE2204/Catch curve/AgeDatWithRecVar.csv", header=T, row.names = NULL)

par(mfrow=c(1,2))
plot(data_var$Age, data_var$Frequency, "o", main="Variable recruitment", cex.main=1.0, 
        xlab="Age class",ylab="Frequency", frame=F, ylim = c(0,130))
plot(data_const$Age, data_const$Frequency, "o", main="Constant recruitment", cex.main=1.0, 
        xlab="Age class",ylab="Frequency", frame=F, ylim = c(0,130))
```

Notice that the variable recruitment frequency plot (on the left) looks quite different to the constant recruitment data we have been using so far (on the right). Most importantly there is a huge spike in frequency at age 11 with around 120 individuals. 

Recruitment spikes might result from a "good year" where currents were favourable and/or spawning was highly successful. 

You might be able to tell already that this is going to complicate our analysis. 

Lets first look at the minimum and maximum cutoffs:

```{r}
MinCutOff = data_var[which.max(data_var$Frequency),"Age"] + 1
MaxCutOff <- data_var[match( 1 , data_var$Frequency), "Age"]
plot(data_var$Age, data_var$Frequency, "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F, ylim=c(0,130))
abline(v = c(MinCutOff, MaxCutOff), col="red", lty=2)

```

Instantly because of this recruitment spike, the minimum cut off has shifted to 12. It was 8 in our last analysis. 

Lets see what happens when we fit the model. 

```{r}
data_analysis<- data_var[ data_var$Age >= MinCutOff & data_var$Age <= MaxCutOff,  ]

CatchCurveModel_var <- lm(log(Frequency) ~ Age, data = data_analysis)
EstlnFreq <- predict(lm(log(Frequency) ~ Age, data = data_analysis))
plot(data_analysis$Age, log(data_analysis$Frequency), "o", main="Age sample data", cex.main=1.0,
        xlab="Age class",ylab="Frequency", frame=F)
lines(data_analysis$Age,EstlnFreq,"l",col="blue")

```

The model does not seem to fit as well as it did for the contant recruitment data. In particular, there seems to be an outlier at age 16. 

```{r}
Z <- CatchCurveModel_var$coefficients[[2]]*-1
Z
F <- Z - M
F
```
Our total mortality rate (Z) is now 0.18. This is much lower than the previous estimate of 0.23. 

This corresponds to a fishing mortality rate of 0.08. 

This is less than the narural mortality rate for WA Dhufish of 0.10, suggesting that we don't think this population is overfished. 


## Some words of caution

Catch curve analysis has a lot of assumptions. These include:

1) Fishing mortality has been constant. Remember we estimated just one number for all fish aged 8 to 25. 

2) Annual recruitment is constant. This was inherent in our first dataset but is violated by our second data set. 

3) At our minimum cut-off of peak frequency + 1 the fish population is fully exposed to the fishery (we say fully recruited)

4) Our sample reflects the underlying age structure of the population

5) All fish are aged correctly - it can be quite tricky to count otolith rings

6) The relationship between maximum age and natural mortlity rate is true for our species

7) The rule of thumb that Fmsy = M is true. This is a very coarse approximation. 


Often in assessing a fish stock a whole range of methods will be used with different sets of assumptions. For those interested, you can see a good example in the 2007 stock assessment for WA Dhufish (which used catch curve analysis amongst other methods) <https://paperpile.com/shared/pj8qFH> 
